---
title: "scRNA-seq Analysis of SARS-CoV-2 Infection"
author: "James Hughes"
date: "2026-03-19"
output:
  html_document:
    df_print: paged
  pdf_document: default
header-includes: \usepackage{float}
---

```{r setup, message=FALSE, warning=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  dev = "png",
  dpi = 150,
  cache = FALSE
)
# Reproducibility
set.seed(1234)
options(scipen = 999)

# Parallel-safe RNG (important with future)
RNGkind("L'Ecuyer-CMRG")

# setwd("C:/Users/james/OneDrive/Desktop/Covid 19/scRNA-seq")

results_dir <- "results"
exports_dir <- "data"

dir.create(results_dir, showWarnings = FALSE)
dir.create(exports_dir, showWarnings = FALSE)

save_plot <- function(plot_obj, filename, width = 7, height = 5, dpi = 200) {
  ggplot2::ggsave(
    filename = file.path(results_dir, filename),
    plot = plot_obj,
    width = width,
    height = height,
    dpi = dpi,
    bg = "white"
  )
}

fmt_int <- function(x) {
  format(as.integer(round(x)), big.mark = ",", scientific = FALSE, trim = TRUE)
}

fmt_num <- function(x, digits = 1) {
  format(round(x, digits), nsmall = digits, scientific = FALSE, trim = TRUE)
}
```

# Introduction

Single-cell RNA sequencing (scRNA-seq) enables transcriptional profiling at the resolution of individual cells, allowing identification of distinct cell populations and their gene expression programmes. In this study, scRNA-seq was used to investigate the peripheral immune response in a publicly available PBMC dataset associated with SARS-CoV-2 infection. The aim was to identify major immune cell populations, define cluster-specific marker genes, and determine which cell types may underlie the immune-related signals observed in bulk RNA-seq studies.

# Methods

## Setup
```{r libraries, message=FALSE, warning=FALSE}
library(Seurat)
library(dplyr)
library(ggplot2)
library(Matrix)
library(patchwork)
library(DropletUtils)
library(tidyverse)
library(SingleR)
library(harmony)
library(future)
library(celldex)

available_workers <- parallel::detectCores(logical = FALSE)
available_workers <- if (is.na(available_workers)) 2 else available_workers
marker_workers <- max(1, min(2, available_workers - 1))
future::plan(sequential)
options(future.globals.maxSize = 8 * 1024^3)
```


### Load Data

```{r load-data}
counts <- Read10X(data.dir = "C:/Users/james/OneDrive/Desktop/Covid 19/scRNA-seq")

dim(counts)

obj <- CreateSeuratObject(
  counts = counts,
  project = "COVID_PBMC_scRNA",
  min.cells = 3,
  min.features = 200
)

obj
dim(obj)

genes_initial <- nrow(obj)
cells_initial <- ncol(obj)
```

### Quality Control

```{r qc}
obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^MT-") ## Calculate % of mitochondrial gene expression
qc_plot <- VlnPlot(
  obj,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  pt.size = 0
) ## Shows distribution for each cell
save_plot(qc_plot, "qc_violin.png", width = 10, height = 4)
qc_plot
```
Quality control (QC) was performed to remove low-quality and artefactual cells from the dataset. Raw scRNA-seq data can include empty droplets, dying cells, and doublets, which may distort downstream analysis. Cells were assessed using the number of detected genes (nFeature_RNA), total RNA counts (nCount_RNA), and the percentage of mitochondrial expression (percent.mt). Low gene counts indicate poor-quality cells, unusually high gene or RNA counts may reflect doublets, and elevated mitochondrial expression is consistent with stressed or dying cells.

### Filtering
```{r filtering}
obj <- subset(obj,
  subset = nFeature_RNA > 300 &
           nFeature_RNA < 6000 &
           nCount_RNA < 40000 &
           percent.mt < 15
)

obj

cells_filtered <- ncol(obj)
genes_filtered <- nrow(obj)
```
Cells were filtered based on quality control metrics to remove low-quality and artefactual observations. Thresholds were applied to retain cells with a sufficient number of detected genes, while excluding cells with unusually high gene or RNA counts indicative of doublets. Additionally, cells with elevated mitochondrial gene expression were removed, as this reflects cellular stress or degradation. This filtering step ensures that downstream analyses are performed on high-quality, biologically meaningful single cells.

### Gene Filtering
```{r gene-filtering}
counts_mat <- GetAssayData(obj, layer = "counts")
keep_genes <- Matrix::rowSums(counts_mat > 0) >= 10
obj <- subset(obj, features = rownames(obj)[keep_genes])

genes_filtered <- nrow(obj)
```
Genes detected in fewer than 10 cells were removed prior to normalisation. This reduces the number of extremely sparse features, which improves memory usage and runtime while retaining the genes that contribute most meaningfully to downstream clustering and marker detection.

### Normalisation
```{r normalization}
obj <- NormalizeData(
  obj,
  normalization.method = "LogNormalize",
  scale.factor = 10000,
  block.size = 2000,
  verbose = FALSE
)
```
The data were normalised to account for differences in sequencing depth across cells, allowing gene expression levels to be compared across the dataset.

### Variable Features
```{r variable-features}
obj <- FindVariableFeatures(obj, selection.method = "vst", nfeatures = 2000)
variable_plot <- VariableFeaturePlot(obj)
save_plot(variable_plot, "variable_features.png", width = 8, height = 5)
variable_plot
```
Highly variable genes were identified to capture the most informative features, focusing downstream analysis on genes with meaningful biological variation.

### Scaling
```{r scaling}
obj <- ScaleData(obj, features = VariableFeatures(obj))
```
The data were scaled so that expression values for each gene were centred and standardised across cells. This step is important prior to principal component analysis, as it prevents genes with high variance or large absolute expression values from disproportionately influencing dimensionality reduction.

### PCA
```{r pca}
obj <- RunPCA(obj, features = VariableFeatures(obj))
elbow_plot <- ElbowPlot(obj) +
  scale_x_continuous(breaks = 1:20) +
  ggtitle("Principal Component Selection (Elbow Plot)") +
    xlab("Principal Component") +
    ylab("Standard Deviation")  +
    theme_classic()

save_plot(elbow_plot, "elbow_plot.png", width = 7, height = 5)
elbow_plot
```   
Principal component analysis (PCA) was performed to reduce dimensionality and summarise the major sources of variation in the dataset. Inspection of the elbow plot indicated that the first six principal components captured the main structured sources of variation in the dataset and were therefore retained for downstream analysis.

### Clustering
```{r clustering}
obj <- FindNeighbors(obj, dims = 1:6)
obj <- FindClusters(obj, resolution = 0.3)
```
Cells were clustered using a graph-based approach implemented in Seurat. A shared nearest neighbour graph was constructed from the selected principal components, and clusters were identified using community detection. This step groups transcriptionally similar cells, allowing putative cell populations or cellular states to be resolved in an unsupervised manner. The clustering resolution was set to 0.3 to generate a moderate level of granularity while avoiding over-fragmentation of the dataset.

### UMAP
```{r umap}
obj <- RunUMAP(obj, dims = 1:6, seed.use = 1234)
umap_clusters <- DimPlot(obj, reduction = "umap", label = TRUE, raster = TRUE)
save_plot(umap_clusters, "umap_clusters.png", width = 8, height = 6)
umap_clusters
```
Uniform Manifold Approximation and Projection (UMAP) was used to visualise the high-dimensional scRNA-seq data in two dimensions. UMAP preserves local structure in the data and allows transcriptionally similar cells to be displayed close together, facilitating interpretation of cluster relationships and population structure.

### Marker Gene Identification
```{r marker-identification}
future::plan(multisession, workers = marker_workers)
markers <- FindAllMarkers(
  obj,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)
future::plan(sequential)

top5_markers <- markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 5, with_ties = FALSE) %>%
  ungroup()

head(markers)
```
Marker genes were identified for each cluster using differential expression analysis implemented in Seurat. For each cluster, expression was compared against all remaining cells to identify genes that were significantly upregulated and characteristic of that population. Only positive markers were retained, and minimum expression and fold-change thresholds were applied to prioritise robust and biologically informative genes.

### Top Marker Genes by Cluster
```{r echo=FALSE}
top5_markers <- markers %>%
  filter(p_val_adj < 0.05) %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 5, with_ties = FALSE) %>%
  ungroup()

knitr::kable(
  top5_markers[, c("cluster", "gene", "avg_log2FC", "pct.1", "p_val_adj")],
  digits = 3,
  caption = "Top five marker genes identified for each cluster."
)
```

### Canonical Immune Marker Expression
```{r canonical-markers}
canonical_dotplot <- DotPlot(
  obj,
   features = c(
    "CD3D", "IL7R",      # T cells
    "CD8A", "CD8B",      # CD8 T cells
    "MS4A1",             # B cells
    "NKG7", "GNLY",      # NK cells
    "LYZ", "S100A8",     # monocytes
    "FCER1A", "CST3",    # dendritic cells
    "PPBP"               # platelets
  )
) + RotatedAxis()

save_plot(canonical_dotplot, "canonical_marker_dotplot.png", width = 10, height = 6)
canonical_dotplot
```
### Cluster Annotation
```{r cluster-annotation}
Idents(obj) <- "seurat_clusters"
levels(Idents(obj))

obj <- RenameIdents(obj,
  `0` = "Cytotoxic lymphocytes",
  `1` = "CD8 T cells",
  `2` = "Classical monocytes",
  `3` = "Activated monocytes",
  `4` = "Naive CD4 T cells",
  `5` = "B cells",
  `6` = "Inflammatory monocytes",
  `7` = "Macrophage-like monocytes",
  `8` = "Endothelial-like cells",
  `9` = "Activated T cells",
  `10` = "Proliferating cells",
  `11` = "Dendritic cells",
  `12` = "Erythrocyte-like cells",
  `13` = "Plasmacytoid dendritic cells"
)

annotated_umap <- DimPlot(
  obj,
  reduction = "umap",
  label = TRUE,
  label.size = 3,
  repel = TRUE,
  raster = TRUE
)
save_plot(annotated_umap, "annotated_umap.png", width = 8, height = 6)
annotated_umap
```
Cluster identities were assigned based on canonical marker gene expression and cluster-specific differentially expressed genes. This enabled biologically meaningful annotation of the unsupervised clusters into major immune cell populations and smaller specialised cell subsets.

### Annotation Validation (SingleR)
```{r singler-validation}

ref <- celldex::HumanPrimaryCellAtlasData()

pred <- SingleR(
  test = GetAssayData(obj, layer = "data"),
  ref = ref,
  labels = ref$label.main
)

obj$SingleR_labels <- pred$labels

obj$SingleR_main <- gsub("_", " ", obj$SingleR_labels)
singler_plot <- DimPlot(obj, group.by = "SingleR_main", label = TRUE, repel = TRUE, raster = TRUE)
save_plot(singler_plot, "singler_umap.png", width = 9, height = 6)
singler_plot
```
SingleR broadly supported the manual annotations at the lineage level. The reference-based labels recovered major B-cell, T-cell/NK-cell, monocyte/macrophage, dendritic-cell, and platelet/erythroblast-like populations in the same UMAP regions identified by marker-guided annotation, increasing confidence that the clustering captures biologically meaningful immune structure rather than technical artefact.


### Cluster Sizes

```{r echo=FALSE}
cluster_sizes <- data.frame(
  Cluster = names(table(Idents(obj))),
  Cells = as.integer(table(Idents(obj))),
  Proportion = sprintf("%.1f%%", 100 * as.integer(table(Idents(obj))) / sum(table(Idents(obj))))
)

cluster_sizes <- cluster_sizes %>%
  arrange(desc(Cells))

total_annotated_cells <- sum(cluster_sizes$Cells)
n_clusters <- nrow(cluster_sizes)
monocyte_cells <- sum(cluster_sizes$Cells[cluster_sizes$Cluster %in% c(
  "Classical monocytes",
  "Activated monocytes",
  "Inflammatory monocytes",
  "Macrophage-like monocytes"
)])
monocyte_pct <- round(100 * monocyte_cells / total_annotated_cells, 1)
bcell_cells <- cluster_sizes$Cells[cluster_sizes$Cluster == "B cells"]
bcell_pct <- round(100 * bcell_cells / total_annotated_cells, 1)
proliferating_cells <- cluster_sizes$Cells[cluster_sizes$Cluster == "Proliferating cells"]
proliferating_pct <- round(100 * proliferating_cells / total_annotated_cells, 1)

knitr::kable(
  cluster_sizes,
  caption = "Cluster sizes after annotation."
)
```

## Results Summary

The dataset contained `r fmt_int(genes_initial)` genes across `r fmt_int(cells_initial)` cells after Seurat object creation. Following quality filtering, `r fmt_int(cells_filtered)` high-quality cells and `r fmt_int(genes_filtered)` genes remained for downstream analysis. Clustering and manual annotation resolved `r n_clusters` transcriptionally distinct populations, with the largest compartments comprising cytotoxic lymphocytes (`r cluster_sizes$Proportion[cluster_sizes$Cluster == "Cytotoxic lymphocytes"]`), CD8 T cells (`r cluster_sizes$Proportion[cluster_sizes$Cluster == "CD8 T cells"]`), classical monocytes (`r cluster_sizes$Proportion[cluster_sizes$Cluster == "Classical monocytes"]`), activated monocytes (`r cluster_sizes$Proportion[cluster_sizes$Cluster == "Activated monocytes"]`), naive CD4 T cells (`r cluster_sizes$Proportion[cluster_sizes$Cluster == "Naive CD4 T cells"]`), and B cells (`r cluster_sizes$Proportion[cluster_sizes$Cluster == "B cells"]`). Together, the four monocyte/macrophage-associated clusters accounted for `r fmt_num(monocyte_pct)`% of all annotated cells, indicating a substantial myeloid component alongside lymphoid populations.

The marker gene analysis strongly supports these annotations. Cytotoxic lymphocytes were characterised by `SH2D1B`, `KLRF1`, and `KLRC2`, while the CD8 T-cell cluster showed strong enrichment for `GZMK`, `CD8A`, and `CD8B`. Classical and activated monocyte populations expressed inflammatory and myeloid markers including `S100A8`, `CXCL2`, `EREG`, `LGALS2`, and `THBD`, whereas B cells were defined by `IGHD`, `CD79A`, `TCL1A`, and `EBF1`. Smaller but biologically informative clusters were also well resolved, including proliferating cells marked by `CDC45`, `PKMYT1`, `MCM10`, `TYMS`, and `CDC6`, dendritic cells marked by `FCER1A`, `CD1C`, and `CLEC10A`, and plasmacytoid dendritic cells marked by `LILRA4` and `CLEC4C`.

The canonical marker dot plot and the SingleR validation step further reinforce the biological interpretation of the dataset. Canonical lineage markers showed the expected patterns, with `MS4A1` enriched in B cells, `CD3D`/`IL7R` and `CD8A`/`CD8B` enriched in T-cell populations, `NKG7`/`GNLY` highlighting cytotoxic lymphocytes, and `LYZ`, `S100A8`, `FCER1A`, and `CST3` distinguishing myeloid and dendritic populations. SingleR recovered broadly concordant B-cell, T-cell/NK-cell, monocyte/macrophage, dendritic-cell, and erythroid/platelet-like labels, suggesting that the manually assigned identities are robust at the lineage level.

## Link to Bulk RNA-seq

This single-cell analysis provides a useful cell-type framework for interpreting the bulk RNA-seq results. In the bulk analysis, 3,879 significant differentially expressed genes were identified, including 3,688 upregulated and 191 downregulated genes, and the strongest enriched pathways included `immunoglobulin mediated immune response`, `B cell mediated immunity`, `antimicrobial humoral response`, and multiple chromosome-segregation and mitotic cell-cycle terms. The scRNA-seq results help localise those whole-sample signals to specific immune populations.

The bulk enrichment for humoral and B-cell-associated pathways is well supported by the presence of a substantial B-cell compartment in the single-cell data (`r fmt_num(bcell_pct)`% of filtered cells), defined by canonical markers such as `CD79A`, `IGHD`, `TCL1A`, and `EBF1`. Likewise, the large monocyte/macrophage component (`r fmt_num(monocyte_pct)`% combined) provides a plausible cellular basis for innate immune activation within the PBMC compartment, while the clear cytotoxic lymphocyte and CD8 T-cell clusters indicate an additional adaptive/cytotoxic axis. Finally, the proliferating-cell cluster, although smaller (`r fmt_num(proliferating_pct)`% of cells), offers a single-cell correlate for the strong cell-cycle and chromosome-segregation signatures seen in the bulk RNA-seq report, alongside highly proliferative bulk markers such as `TK1`, `PLK1`, `RRM2`, `UBE2C`, and `CCNA2`.

Taken together, the bulk RNA-seq analysis identifies the dominant transcriptional programmes present at whole-sample level, whereas the scRNA-seq analysis shows which PBMC populations are available to generate those signatures. The two analyses are therefore complementary: bulk RNA-seq provides sensitivity for pathway-level shifts across the cohort, while scRNA-seq provides the cellular resolution needed to interpret where immune activation, humoral responses, and proliferative signals are likely originating.

## Limitations

Although the current workflow provides a biologically meaningful overview of PBMC cell populations, it does not yet test differential abundance or differential expression between infected and control donors at cell-type level. Additional sample-level metadata and a donor-aware statistical design would be required to determine whether particular populations expand, contract, or undergo transcriptional rewiring in response to SARS-CoV-2 infection. It is also important to note that manual annotation remains an interpretation step, particularly for the smaller endothelial-like and erythrocyte-like clusters, even though the marker genes and SingleR labels broadly support the major lineage assignments. The current analysis is therefore best interpreted as a strong descriptive single-cell atlas and annotation workflow rather than a full comparative disease-model study.

## Exported Outputs

```{r}
write.csv(markers, file.path(exports_dir, "all_markers.csv"), row.names = FALSE)
write.csv(top5_markers, file.path(exports_dir, "top5_markers_by_cluster.csv"), row.names = FALSE)
write.csv(cluster_sizes, file.path(exports_dir, "cluster_sizes.csv"), row.names = FALSE)
write.csv(
  data.frame(Cell = colnames(obj), Annotation = obj$SingleR_main),
  file.path(exports_dir, "singleR_labels.csv"),
  row.names = FALSE
)
```

The main tables and annotation outputs were exported to the `data` folder, and key figures were saved to the `results` folder to make the project easier to inspect and present on GitHub.

## Conclusion

Overall, this scRNA-seq analysis resolves the PBMC dataset into `r n_clusters` annotated cell populations spanning lymphoid, myeloid, dendritic, proliferative, and minor specialised compartments. The results show that the dataset contains a strong mixture of cytotoxic lymphocytes, CD8 T cells, monocytes, naive CD4 T cells, and B cells, supported by cluster-specific markers and broadly concordant SingleR assignments. When interpreted alongside the bulk RNA-seq analysis, these data suggest that the immune and humoral programmes observed at whole-sample level are likely driven by a combination of B-cell-rich, monocyte-rich, and cytotoxic lymphocyte populations, with a smaller proliferative compartment contributing to the cell-cycle signature. This makes the scRNA-seq workflow a strong single-cell complement to the bulk RNA-seq study and a useful extension of the overall COVID-19 transcriptomics project.

## Session Information

```{r session-info}
sessionInfo()
```
