---
title: "RNA-seq Analysis of SARS-CoV-2 Infection"
author: "James Hughes"
date: "2026-09-05"
output:
  html_document:
    df_print: paged
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 7, fig.height = 6)
set.seed(42)

repo_root <- if (file.exists(file.path("data", "GSE152418_counts.txt"))) {
  normalizePath(".", mustWork = TRUE)
} else {
  normalizePath("..", mustWork = TRUE)
}
knitr::opts_knit$set(root.dir = repo_root)

counts_path <- file.path(repo_root, "data", "GSE152418_counts.txt")
metadata_path <- file.path(repo_root, "data", "metadata.csv")
tables_dir <- file.path(repo_root, "outputs", "tables")
figures_dir <- file.path(repo_root, "outputs", "figures")
dir.create(tables_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figures_dir, recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(fig.path = paste0(figures_dir, "/"))

normalise_sample_ids <- function(x) gsub("[-.]", "_", trimws(x))

validate_bulk_inputs <- function(expr, meta) {
  expr <- as.matrix(expr)
  if (!is.numeric(expr) || any(!is.finite(expr)) || any(expr < 0) || any(expr != floor(expr))) {
    stop("Counts must be finite nonnegative integers.")
  }
  ids <- normalise_sample_ids(colnames(expr))
  if (is.null(rownames(expr)) || anyNA(rownames(expr)) || any(!nzchar(rownames(expr))) || anyDuplicated(rownames(expr))) {
    stop("Gene IDs must be present and unique.")
  }
  if (is.null(ids) || anyNA(ids) || any(!nzchar(ids)) || anyDuplicated(ids)) {
    stop("Sample IDs are missing or collide after normalization.")
  }
  if (!all(c("sample", "condition") %in% names(meta))) {
    stop("Metadata require sample and condition columns.")
  }
  meta$sample <- normalise_sample_ids(meta$sample)
  if (anyNA(meta$sample) || any(!nzchar(meta$sample)) || anyDuplicated(meta$sample) || !setequal(ids, meta$sample)) {
    stop("Metadata and count samples must match exactly.")
  }
  meta <- meta[match(ids, meta$sample), , drop = FALSE]
  if (anyNA(meta$condition) || any(!meta$condition %in% c("Control", "Infected"))) {
    stop("Unknown condition in metadata.")
  }
  meta$condition <- factor(meta$condition, levels = c("Control", "Infected"))
  if (any(table(meta$condition) < 2L)) stop("At least two libraries per condition are required.")
  if (any(colSums(expr) == 0)) stop("Empty count library.")
  rownames(meta) <- ids
  colnames(expr) <- ids
  list(counts = expr, metadata = meta)
}
```

# Introduction

This analysis describes bulk PBMC expression in GSE152418 using the original 34-library cohort and a condition-only DESeq2 design. The historical `Infected` label contains 16 acute COVID-19 libraries and one convalescent library. Repeat-draw identifiers are also present, so donor independence is unresolved. The workflow is therefore treated as an exploratory reproduction rather than a definitive disease-effect analysis.

Counts are gene-level inputs rather than FASTQ reads. Bulk expression cannot establish cell-intrinsic regulation, pathway activity, or causality.

# Methods

## Setup

```{r setup}
library(apeglm)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(DESeq2)
library(enrichplot)
library(clusterProfiler)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(dplyr)
library(tibble)
BiocParallel::register(BiocParallel::SerialParam())
```

## Load and validate data

```{r load-data}
expr <- read.delim(counts_path, row.names = 1, check.names = FALSE)
colnames(expr) <- normalise_sample_ids(colnames(expr))
meta <- read.csv(metadata_path, check.names = FALSE, stringsAsFactors = FALSE)
validated <- validate_bulk_inputs(expr, meta)
expr <- validated$counts
meta <- validated$metadata
sample_counts <- table(meta$condition)
```

Metadata are matched one-to-one to normalized count-column identifiers before analysis. Collisions, omissions, unexpected labels, invalid counts, duplicate gene IDs, and empty libraries cause the analysis to stop.

```{r sample-summary, echo=FALSE}
knitr::kable(as.data.frame(sample_counts), col.names = c("Condition", "Number of libraries"), caption = "Sample composition of the bulk RNA-seq dataset.")
```

## Pre-processing and differential expression

Genes with fewer than 10 total counts across all libraries are removed. Differential expression is estimated with DESeq2 and shrunken with apeglm.

```{r differential-expression}
expr <- expr[, rownames(meta)]
genes_before_filter <- nrow(expr)
expr <- expr[rowSums(expr) >= 10, ]
genes_after_filter <- nrow(expr)

dds <- DESeqDataSetFromMatrix(countData = expr, colData = meta, design = ~ condition)
dds <- DESeq(dds)
res <- lfcShrink(dds, coef = "condition_Infected_vs_Control", type = "apeglm")

res_df <- as.data.frame(res) |>
  tibble::rownames_to_column("gene_id") |>
  dplyr::mutate(ensembl_id = gsub("\\..*", "", gene_id))

symbol_candidates <- AnnotationDbi::mapIds(
  org.Hs.eg.db,
  keys = unique(res_df$ensembl_id),
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "list"
)
symbol_candidates <- lapply(symbol_candidates, function(x) sort(unique(x[!is.na(x)])))
res_df$symbol <- vapply(res_df$ensembl_id, function(id) {
  values <- symbol_candidates[[id]]
  if (length(values) == 1L) values else id
}, character(1))

res_df <- res_df |>
  dplyr::mutate(
    deg_class = dplyr::case_when(
      is.na(padj) ~ "Not tested",
      padj < 0.05 & log2FoldChange > 1 ~ "Up",
      padj < 0.05 & log2FoldChange < -1 ~ "Down",
      TRUE ~ "NS"
    ),
    significant = ifelse(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1, "Significant", "NS")
  )
```

After filtering, `r genes_after_filter` genes were retained from `r genes_before_filter`. Differentially expressed genes are called at adjusted p-value < 0.05 and absolute shrunken log2 fold change > 1.

# Visualisation

## PCA

```{r pca}
vsd <- vst(dds, blind = FALSE)
pca_plot <- plotPCA(vsd, intgroup = "condition") + labs(title = "PCA of Gene Expression")
print(pca_plot)
ggsave(file.path(figures_dir, "pca.png"), pca_plot, width = 7, height = 5, dpi = 300)
```

## Volcano plot

```{r volcano}
top_genes <- res_df |> dplyr::filter(!is.na(padj)) |> dplyr::arrange(padj) |> dplyr::slice_head(n = 10) |> dplyr::pull(gene_id)
label_data <- res_df |> dplyr::filter(gene_id %in% top_genes)

volcano_plot <- ggplot(res_df, aes(x = log2FoldChange, y = -log10(pmax(padj, .Machine$double.xmin)), colour = deg_class)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_text_repel(data = label_data, aes(label = symbol), size = 3, max.overlaps = Inf, show.legend = FALSE) +
  scale_colour_manual(values = c("Down" = "navy", "Up" = "firebrick3", "NS" = "gray", "Not tested" = "grey80")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  labs(title = "Volcano Plot of Differential Expression", x = "Shrunken log2 fold change", y = "-log10 adjusted p-value") +
  theme_minimal()
print(volcano_plot)
ggsave(file.path(figures_dir, "volcano.png"), volcano_plot, width = 7, height = 6, dpi = 300)
```

## MA plot

```{r ma-plot}
ma_plot <- ggplot(res_df, aes(x = baseMean, y = log2FoldChange, colour = significant)) +
  geom_point(alpha = 0.3, size = 0.9) +
  scale_x_log10() +
  scale_colour_manual(values = c("NS" = "grey", "Significant" = "red")) +
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = c(-1, 1), linetype = "dashed") +
  labs(title = "MA Plot of Differential Expression", x = "Mean normalized count", y = "Shrunken log2 fold change") +
  theme_minimal()
print(ma_plot)
ggsave(file.path(figures_dir, "ma_plot.png"), ma_plot, width = 7, height = 6, dpi = 300)
```

## Heatmaps

```{r heatmaps}
top_genes_20 <- res_df |> dplyr::filter(!is.na(padj)) |> dplyr::arrange(padj) |> dplyr::slice_head(n = 20) |> dplyr::pull(gene_id)
gene_symbols <- res_df$symbol[match(top_genes_20, res_df$gene_id)]
heatmap_data <- assay(vsd)[top_genes_20, ]
rownames(heatmap_data) <- gene_symbols
annotation_col <- data.frame(condition = meta$condition)
rownames(annotation_col) <- rownames(meta)

png(file.path(figures_dir, "heatmap_top_genes.png"), width = 1800, height = 1500, res = 250)
pheatmap(heatmap_data, scale = "row", color = colorRampPalette(c("navy", "white", "firebrick3"))(50), annotation_col = annotation_col, show_colnames = FALSE, fontsize_row = 8)
dev.off()
pheatmap(heatmap_data, scale = "row", color = colorRampPalette(c("navy", "white", "firebrick3"))(50), annotation_col = annotation_col, show_colnames = FALSE, fontsize_row = 8)

sample_dists <- dist(t(assay(vsd)))
png(file.path(figures_dir, "sample_distance_heatmap.png"), width = 1800, height = 1500, res = 250)
pheatmap(as.matrix(sample_dists), annotation_col = annotation_col, color = colorRampPalette(c("navy", "white", "firebrick3"))(50), show_colnames = FALSE, show_rownames = FALSE)
dev.off()
pheatmap(as.matrix(sample_dists), annotation_col = annotation_col, color = colorRampPalette(c("navy", "white", "firebrick3"))(50), show_colnames = FALSE, show_rownames = FALSE)
```

# GO gene-set enrichment analysis

```{r gsea}
rank_input <- res_df |> dplyr::filter(is.finite(log2FoldChange))
gene_df <- AnnotationDbi::select(org.Hs.eg.db, keys = unique(rank_input$ensembl_id), keytype = "ENSEMBL", columns = "ENTREZID") |>
  dplyr::distinct(ENSEMBL, ENTREZID) |>
  dplyr::group_by(ENSEMBL) |>
  dplyr::mutate(n_entrez = dplyr::n_distinct(ENTREZID, na.rm = TRUE)) |>
  dplyr::ungroup() |>
  dplyr::left_join(rank_input[, c("ensembl_id", "log2FoldChange")], by = c("ENSEMBL" = "ensembl_id")) |>
  dplyr::arrange(ENTREZID, dplyr::desc(abs(log2FoldChange)), ENSEMBL) |>
  dplyr::group_by(ENTREZID) |>
  dplyr::mutate(selected = !is.na(ENTREZID) & n_entrez == 1L & cumsum(n_entrez == 1L) == 1L) |>
  dplyr::ungroup()

ranked <- gene_df |> dplyr::filter(selected) |> dplyr::arrange(dplyr::desc(log2FoldChange), ENTREZID)
gene_list <- setNames(ranked$log2FoldChange, ranked$ENTREZID)
stopifnot(!anyDuplicated(names(gene_list)), all(is.finite(gene_list)))
set.seed(42)
ego_gsea <- gseGO(geneList = gene_list, OrgDb = org.Hs.eg.db, keyType = "ENTREZID", ont = "BP", seed = TRUE, BPPARAM = BiocParallel::SerialParam(), verbose = FALSE)

if (nrow(as.data.frame(ego_gsea)) > 0) {
  gsea_dotplot <- dotplot(ego_gsea, showCategory = 8)
  print(gsea_dotplot)
  ggsave(file.path(figures_dir, "gsea_dotplot.png"), gsea_dotplot, width = 7, height = 6, dpi = 300)
  png(file.path(figures_dir, "gsea_enrichment.png"), width = 1800, height = 1500, res = 250)
  print(gseaplot2(ego_gsea, geneSetID = 1))
  dev.off()
  gseaplot2(ego_gsea, geneSetID = 1)
}
```

# Results

```{r result-summary}
sig <- res_df |> dplyr::filter(!is.na(padj), padj < 0.05, abs(log2FoldChange) > 1)
n_sig <- nrow(sig)
n_up <- sum(sig$log2FoldChange > 0)
n_down <- sum(sig$log2FoldChange < 0)
top <- sig |> dplyr::arrange(padj) |> dplyr::slice_head(n = 5)
top_gene_text <- paste(top$symbol, collapse = ", ")
```

The model identifies `r n_sig` significant genes, including `r n_up` higher and `r n_down` lower in the historical `Infected` group. Top-ranked significant genes include `r top_gene_text`.

```{r result-tables, echo=FALSE}
knitr::kable(data.frame(Metric = c("Significant DEGs", "Higher in Infected", "Lower in Infected"), Count = c(n_sig, n_up, n_down)), caption = "Differential-expression summary.")

knitr::kable(top |> dplyr::select(symbol, log2FoldChange, padj), digits = 3, col.names = c("Gene", "Log2 fold change", "Adjusted p-value"), caption = "Top five differentially expressed genes.")

if (nrow(as.data.frame(ego_gsea)) > 0) {
  gsea_display <- head(ego_gsea@result[, c("Description", "NES", "p.adjust")])
  colnames(gsea_display) <- c("Pathway", "NES", "Adjusted p-value")
  knitr::kable(gsea_display, digits = 3, caption = "Top GO Biological Process terms identified by GSEA.")
}
```

# Interpretation and limitations

These results are expression associations under the specified condition-only model. The cohort mixes acute infection and convalescence, and repeat-draw identifiers mean donor independence is unresolved. Severity, sex, timing, and other covariates are not modelled. Bulk expression also cannot distinguish altered cell proportions from within-cell expression changes.

# Export results

```{r exports}
write.csv(res_df, file.path(tables_dir, "all_DESeq2_results_with_symbols.csv"), row.names = FALSE)
write.csv(sig, file.path(tables_dir, "all_significant_DEGs.csv"), row.names = FALSE)
write.csv(sig |> dplyr::arrange(padj) |> dplyr::slice_head(n = 100), file.path(tables_dir, "top_100_DEGs.csv"), row.names = FALSE)
write.csv(as.data.frame(ego_gsea), file.path(tables_dir, "gsea_results.csv"), row.names = FALSE)
write.csv(meta, file.path(tables_dir, "analysis_metadata.csv"), row.names = FALSE)
write.csv(gene_df, file.path(tables_dir, "gsea_mapping_audit.csv"), row.names = FALSE)
write.csv(ranked, file.path(tables_dir, "gsea_ranked_genes.csv"), row.names = FALSE)
writeLines(capture.output(sessionInfo()), file.path(tables_dir, "session_info.txt"))
```

# References

- GSE152418: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE152418
- Love MI, Huber W, Anders S. DESeq2. Genome Biology (2014), doi:10.1186/s13059-014-0550-8.
- Yu G et al. clusterProfiler. OMICS (2012), doi:10.1089/omi.2011.0118.

```{r session-info}
sessionInfo()
```
